---
title: "02-Logistic-regression"
output: html_document
---



### Quick tasks 

1. The `esoph` data includes counts of a case-control study of esophageal cancer, according to age (`agegp`), alcohol consumption (`alcgp`) and tobacco consumption (`tobgp`). Fit a logistic regression model to the responses, formed by both case and control counts, using all covariates available. Print a summary of the model fit, and also the ANOVA table.

```{r, echo = FALSE, eval = FALSE}
myresp <- cbind(esoph$ncases, esoph$ncontrols)
eso.logit <- glm(myresp ~ esoph$agegp + esoph$alcgp + esoph$tobgp, family = binomial)
summary(eso.logit)
anova(eso.logit)
```

2. Fit also a model which includes also an interaction effect between alcohol and tobacco consumption. Print again the model summary and the anova table of the model fit.

```{r, echo = FALSE, eval = FALSE}
myresp <- cbind(esoph$ncases, esoph$ncontrols)
eso.logit.i <- glm(myresp ~ esoph$agegp + esoph$alcgp * esoph$tobgp, family = binomial)
summary(eso.logit.i)
anova(eso.logit.i)
```
