---
title: "01-Introduction-to probabilistic-distributions"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = param, eval= param, include=T)
```

A student in statistics wants to make a small experiment. He throws 50 times a coin and counts the number of times he gets head. Compute the probability to obtain head 20, 25 and 35 times using the function `dbinom`. Plot the probability plot resulting from this experiment.

```{r}
dbinom(c(20,25,35), size = 50, prob =0.5)
plot(c(0:50),dbinom(c(0:50), size = 50, prob =0.5), ylim=c(0,0.2),col="blue",ylab="p")
```

500 students from the statistical school throw 5 times a coin and report the average number heads obtained per throw. Simulate the results of the 500 students using the `sapply` and `rbinom` function. Plot the histogram of the 500 results obtained.

```{r}
set.seed(42)
means <- NULL
for (k in 1:500){
sample <- rbinom(5, size = 1, prob =0.5)
means <- c(means, mean(sample))
}
hist(means,col="blue")
```

Not satisfied by these results, the students decides to throw the coin 50 times instead of 5. Simulate and plot the results.

```{r}
set.seed(42)
means<-NULL
for (k in 1:500){
sample <- rbinom(50, size = 1, prob =0.5)
means <- c(means, mean(sample))
}
hist(means,col="blue")
```