---
title: "01-comparing-2-groups"
output: html_document
---


### Quick Task

Perform the Student's t test and the Wilcoxon rank-sum test to compare the performances of athletes on the 100m event between the competitions.  

```{r}
library(factoextra)
data(decathlon2)
t.test(X100m ~ Competition, data = decathlon2)
wilcox.test(X100m ~ Competition, data = decathlon2)
```

Now compute for each events and the total number of points between both events in the `decathlon2` dataset the p-values obtained by the student's t test and the wilcoxon test. Display the p-values of both tests for each variable in a data frame. Do you see any differences between the tests?

*Hint : you can create a function that will compute both p-values for one event and use the function `sapply` to apply it to all events.*

```{r}

events <- colnames(decathlon2)[c(1:10,12)]

two.test <- function(response, covariate, dataset){
    formula <- formula(paste0(response, ' ~ ', covariate))
    ttest <- t.test(formula, data = dataset)
    wilcoxtest <- wilcox.test(formula, data = dataset, exact = F)
    pvalues <- c(ttest$p.value, wilcoxtest$p.value)
}

testresults <- sapply(events, two.test, covariate = '	Competition', dataset= decathlon2)

dataTests <- data.frame(ttest = testresults[1,], wtest = testresults[2,]) 
```
