---
title: "02-Testing independence in 2x2 tables"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = param, eval= param, include=T)
```


### Quick tasks

1. Use both the Fisher's exact test as well as the chi-square test to decide if there is evidence of association between learning speed and sex in the `quine` data. Extract the p-value of each test, and add the p-value to the barplot of the data.

```{r}
fisher.test(quine$Lrn, quine$Sex)
myf.test <- fisher.test(quine$Lrn, quine$Sex)
mych.test <- chisq.test(quine$Lrn, quine$Sex)
```


```{r}
plot(quine$Lrn ~ quine$Sex, main = paste("Chi-square p=", round(mych.test$p.value, 2), 
                                         "Fisher ex p=", round(myf.test$p.value, 2)))

```


2. Now consider two other datasets involving variables `X` and `Y`, with the cross-tabulations of the variables given below - first for study 1:

```{r def.n.workex.indep, echo=F, message=FALSE, warning=FALSE}
set.seed(55498)
n <- 100
x1 <- rbinom(n, prob = 0.3, size = 1)
y1 <- rbinom(n, prob = 0.2, size = 1)

x2 <- x1
y2 <- rbinom(n, prob = 0.8*x2, size = 1)

```


```{r,echo=T, eval=T}
table(x1, y1)
```

Now for study 2:

```{r, echo=T, eval=T}
table(x2, y2)
```

Try to use both the chi-square test and Fisher's exact test to investigate whether or not the data supports the null hypothesis of independence between `X` and `Y`.

```{r}
myt1 <- table(x1, y1)
chisq.test(myt1)
fisher.test(myt1)

myt2 <- table(x2, y2)
chisq.test(myt2)
fisher.test(myt2)
```
